Delta-Complete Linear Programming

### 2025-06-09 docs/paper/main.tex
feat: add tuning

The $n$-dimensional Gaussian measure with mean $\mu\in\mathbb{R}^n$ and covariance matrix $\Sigma\in\mathbb{R}^{n\times n}$ is given by
\begin{equation*}
  \mathcal N(dx|\mu, \Sigma) := \frac{dx}{ \sqrt{(2\pi)^n \left|\Sigma\right| }}\exp\left[-\frac{1}{2}(x-\mu)\T\Sigma^{-1\!}(x-\mu)\right],
\end{equation*}
where $|\Sigma|$ is the determinant of $\Sigma$.
The Dirac delta measure $\delta_a:\borel{\X}\rightarrow [0,1]$ concentrated at a point $a\in\X$ is defined as $\delta_a(A)=1$ if $a\in A$ and $\delta_a(A)=0$ otherwise, for any measurable set $A\in\borel{\X}$.
We denote the uniform distribution over $\X$ as $\mathcal{U}_\X$ with realizations $x\sim\mathcal{U}_\X(\cdotx)$.

@@

@@

Substituting the barrier $B\in\Hilbert_M$ of the form \eqref{eq:representer-form-spectral} and the approximation of the CME term in \eqref{eq:cme-term-spectral-approx} into the SIP \eqref{eq:semiinf-prog}, we obtain the \emph{semi-infinite linear program} (SILP)
\begin{subequations}
\begin{align}
    \min_{b, c, \eta, \gamma} \; &\frac{\eta + cT}{\gamma},\label{eq:semiinfLP-prog-objective}\\
    \text{s.t.}\quad
    &\forall x_0\in\X_0:\,\phi_M(x_0)\T b\leq\eta,\label{eq:semiinfLP-prog-initial}\\
    &\forall x_u\in\X_u:\,\phi_M(x_u)\T b\geq\gamma,\label{eq:semiinfLP-prog-unsafe}\\
    &\forall x\in\X: \, \phi_M(x)\T (H b - b) \leq c - \varepsilon\bar{B}\kappa,\label{eq:semiinfLP-prog-kushner}\\
    & \forall x\in\X:\,\phi_M(x)\T b\geq 0,\label{eq:semiinfLP-prog-positive}\\
    &c\geq 0,\,\gamma>\eta\geq 0,\,b\in\R^{2M+1},\nonumber
\end{align}\label{eq:semiinfLP-prog}%
\end{subequations}
with coefficients $\kappa\geq\sigma_f$ and $\bar{B}\geq\norm{b}_2$.

@@

Via Lemma~\ref{lem:trig-bound}, we obtain a finite-constraint relaxation of the SIP \eqref{eq:semiinf-prog} based on the constraint-tightening reasoning
\begin{equation*}
    \frac{1}{2}\big(\mathrm{max}_{\hat{N}}^\X + \mathrm{min}_{\hat{N}}^\X - C_{\hat{N}}\left(\mathrm{max}_{\hat{N}}^\X-\mathrm{min}_{\hat{N}}^\X\right)\big)\geq \mathrm{lb}
    \quad\Longrightarrow\quad
    \forall x\in\X: f(x) \geq \mathrm{lb},
\end{equation*}
where $\mathrm{max}_{\hat{N}}^\X:=\max_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ and $\mathrm{min}_{\hat{N}}^\X:=\min_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ for some desired lower bound $\mathrm{lb}\in\R$.
Thus, we obtain strengthened constraints of the form 
\begin{align*}
    &\frac{1}{2}\big(\mathrm{max}_{\hat{N}}^\X + \mathrm{min}_{\hat{N}}^\X - C_{\hat{N}}\left(\mathrm{max}_{\hat{N}}^\X-\mathrm{min}_{\hat{N}}^\X\right)\big) \geq \mathrm{lb},\\
    \Leftrightarrow\hspace{10pt} &
    \mathrm{min}_{\hat{N}}^\X \geq \frac{2}{C_{\hat{N}}+1}\mathrm{lb} + \frac{C_{\hat{N}}-1}{C_{\hat{N}}+1}\mathrm{max}_{\hat{N}}^\X.
\end{align*}
For an upper bound $\forall x\in\X: f(x) \leq \mathrm{ub}$, we obtain analogous constraints of the form
\begin{equation*}
    \mathrm{max}_{\hat{N}}^\X \geq \frac{2}{C_{\hat{N}}+1}\mathrm{ub} + \frac{C_{\hat{N}}-1}{C_{\hat{N}}+1}\mathrm{min}_{\hat{N}}^\X.
\end{equation*}

\vfill

>>>>

The $n$-dimensional Gaussian measure with mean $\mu\in\mathbb{R}^n$ and covariance matrix $\Sigma\in\mathbb{R}^{n\times n}$ is given by
\begin{equation}
  \label{eq:gaussian-kernel}
  \mathcal N(dx|\mu, \Sigma) := \frac{dx}{ \sqrt{(2\pi)^n \left|\Sigma\right| }}\exp\left[-\frac{1}{2}(x-\mu)\T\Sigma^{-1\!}(x-\mu)\right],
\end{equation}
where $|\Sigma|$ is the determinant of $\Sigma$.
The Dirac delta measure $\delta_a:\borel{\X}\rightarrow [0,1]$ concentrated at a point $a\in\X$ is defined as $\delta_a(A)=1$ if $a\in A$ and $\delta_a(A)=0$ otherwise, for any measurable set $A\in\borel{\X}$.
We denote the uniform distribution over $\X$ as $\mathcal{U}_\X$ with realizations $x\sim\mathcal{U}_\X(\cdotx)$.

@@

A symmetric function $k_\X:\X\times\X\rightarrow\R$ is called a (positive definite) \emph{kernel} (note the distinction from \emph{probability kernels}) if for all $N\in\N_{>0}$ we have $\sum_{i=1}^{N}\sum_{j=1}^{N}a_i a_j\allowbreak k_\X(x_i,x_j) \geq 0$ for $x_1,\ldots,x_N\in\X\subset{\R^n}$ and $a_1,\ldots,a_N\in\R$.
A prominent example is the \emph{squared exponential} (SQExp) kernel \cite{Rasmussen2005GP,Kanagawa2018GPvsKernel}:
\begin{equation}
  \label{eq:gaussian-kernel}
  k_\X(x,x') := \sigma_f^2 \exp\left( -\frac{1}{2} (x-x')\T \Sigma\, (x-x') \right),\quad \Sigma:=\diag(\sigma_l)^{-2},
\end{equation}
with amplitude $\sigma_f^2\geq0$ and lengthscale coefficients $\sigma_l\in\R^n$.
%In the following, we refer to a positive definite kernel as \emph{kernel}, for simplicity.
For this work, we assume that all kernels are bounded on their domain, i.e., $\E_{}[k_\X(x,x)]<\infty$, $x\in\X$.
Given a kernel $k_\X$ on a non-empty set $\X$, there exists a unique corresponding \emph{reproducing kernel Hilbert space} (RKHS) $\Hilbert_{k_\X}$ of functions $f:\X\rightarrow\R$ equipped with an inner product $\innerH{\cdotx}{\cdotx}{\Hilbert_{k_\X}}$ possessing the \emph{reproducing property} such that for any function $f\in\Hilbert_{k_\X}$ and $x\in\X$ we have $f(x)=\innerH{f}{k_\X(\cdotx,x)}{\Hilbert_{k_\X}}$.
Note that $k_\X(\cdotx,x):\X\rightarrow\Hilbert_{k_\X}$ is a real-valued function,
\new{which is} also called an implicit \emph{canonical} \emph{embedding} or \emph{feature map} $\phi_\X$
such that $k_\X(x,x')=\innerH{\phi_\X(x)}{\phi_\X(x')}{\Hilbert_{k_\X}}$ for all $x,x'\in\X$.
For an RKHS $\Hilbert_{k_\X}$, we use the associated feature map $\phi_\X$ and kernel $k_\X$ interchangeably for ease of notation and comprehensibility.
%We will call the kernel associated to an RKHS $\Hilbert_{k_\X}$ the \emph{characteristic} kernel of $\Hilbert_{k_\X}$ and use $\phi_\X$ and $k_\X$ interchangeably for ease of notation and comprehensibility.
%and an implicit \emph{canonical} \emph{embedding} or \emph{feature map} $\phi_\X:\X\rightarrow\Hilbert_{k_\X}$ with an inner product on $\Hilbert_{k_\X}$, i.e., $k_\X(x,x')=\innerH{\phi_\X(x)}{\phi_\X(x')}{\Hilbert_{k_\X}}$, 
%Furthermore, 
The inner product induces the norm $\norm{f}_{\Hilbert_{k_\X}}\!\!\!\!:=\!\!\sqrt{\smash[b]{\innerH{f}{f}{\Hilbert_{k_\X}}}}$ of the RKHS.
%Note that $\Hilbert_{k_\X}$ is a space of functions $f:\X\rightarrow\R$. 
%It is equipped with the infamous \emph{reproducing property} such that for any function $f\in\Hilbert_{k_\X}$ and $x\in\X$ we have $f(x)=\innerH{f}{\phi_\X(x)}{\Hilbert_{k_\X}}$.
%Furthermore, the norm associated with the RKHS $\Hilbert_{k_\X}$ is defined as $\norm{f}_{\Hilbert_{k_\X}}:=\sqrt{\innerH{f}{f}{\Hilbert_{k_\X}}}$.
%We refer the interested reader to XX for a comprehensive dissemination on RKHS.
Throughout this paper, we assume that all RKHSs are \emph{separable}.
Given $N$ i.i.d. samples $\hat X_N:=[\hat{x}_i]_{i=1}^N$ with $\hat{x}_i\in\X$, the
\emph{Gram matrix} of $k_\X$ is given by
$K_{\hat{X}}^N:=[k_\X(\hat{x}_i,\hat{x}_j)]_{i,j=1}^N.$
Furthermore, we define the vector-valued function
$k_{\hat{X}}^N(x)  := [k_\X(x,\hat{x}_i)]_{i=1}^N.$

@@

Substituting the barrier $B\in\Hilbert_M$ of the form \eqref{eq:representer-form-spectral} and the approximation of the CME term in \eqref{eq:cme-term-spectral-approx} into the SIP \eqref{eq:semiinf-prog}, we obtain the \emph{semi-infinite linear program} (SILP)
\begin{subequations}
  \begin{align}
    \min_{b, c, \eta, \gamma} \; & \frac{\eta + cT}{\gamma},\label{eq:semiinfLP-prog-objective}                                                 \\
    \text{s.t.}\quad
                                 & \forall x_0\in\X_0:\,\phi_M(x_0)\T b\leq\eta,\label{eq:semiinfLP-prog-initial}                               \\
                                 & \forall x_u\in\X_u:\,\phi_M(x_u)\T b\geq\gamma,\label{eq:semiinfLP-prog-unsafe}                              \\
                                 & \forall x\in\X: \, \phi_M(x)\T (H b - b) \leq c - \varepsilon\bar{B}\kappa,\label{eq:semiinfLP-prog-kushner} \\
                                 & \forall x\in\X:\,\phi_M(x)\T b\geq 0,\label{eq:semiinfLP-prog-positive}                                      \\
                                 & c\geq 0,\,\gamma>\eta\geq 0,\,b\in\R^{2M+1},\nonumber
  \end{align}\label{eq:semiinfLP-prog}%
\end{subequations}
with coefficients $\kappa\geq\sigma_f$ and $\bar{B}\geq\norm{b}_2$.

@@

Via Lemma~\ref{lem:trig-bound}, we obtain a finite-constraint relaxation of the SIP \eqref{eq:semiinf-prog} based on the constraint-tightening reasoning
\begin{equation*}
  \frac{1}{2}\big(\mathrm{max}_{\hat{N}}^\X + \mathrm{min}_{\hat{N}}^\X - C_{\hat{N}}\left(\mathrm{max}_{\hat{N}}^\X-\mathrm{min}_{\hat{N}}^\X\right)\big)\geq \mathrm{lb}
  \quad\Longrightarrow\quad
  \forall x\in\X: f(x) \geq \mathrm{lb},
\end{equation*}
where $\mathrm{max}_{\hat{N}}^\X:=\max_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ and $\mathrm{min}_{\hat{N}}^\X:=\min_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ for some desired lower bound $\mathrm{lb}\in\R$.
Thus, we obtain strengthened constraints of the form
\begin{align*}
                               & \frac{1}{2}\big(\mathrm{max}_{\hat{N}}^\X + \mathrm{min}_{\hat{N}}^\X - C_{\hat{N}}\left(\mathrm{max}_{\hat{N}}^\X-\mathrm{min}_{\hat{N}}^\X\right)\big) \geq \mathrm{lb}, \\
  \Leftrightarrow\hspace{10pt} &
  \mathrm{min}_{\hat{N}}^\X \geq \frac{2}{C_{\hat{N}}+1}\mathrm{lb} + \frac{C_{\hat{N}}-1}{C_{\hat{N}}+1}\mathrm{max}_{\hat{N}}^\X.
\end{align*}
For an upper bound $\forall x\in\X: f(x) \leq \mathrm{ub}$, we obtain analogous constraints of the form
\begin{equation*}
  \mathrm{max}_{\hat{N}}^\X \geq \frac{2}{C_{\hat{N}}+1}\mathrm{ub} + \frac{C_{\hat{N}}-1}{C_{\hat{N}}+1}\mathrm{min}_{\hat{N}}^\X.
\end{equation*}

\subsection{Tuning the Hyperparameters}
\label{sec:tuning-hyperparameters}

Kernel methods do not require the expensive learning process of other machine learning methods, such as neural networks.
However, they still depend on a number of \hp, such as the kernel bandwidth $\sigma_f$ and the number of Fourier coefficients $M$.
Changes in their values can have a significant impact on the estimator's efficiency and accuracy.
The process of finding good values for these \hp is known as \emph{hyperparameter tuning}, and it is problem-dependent.

There are several approaches to find the optimal \hp for a given problem.
A cheap option which can be used as a starting point when working with a Gaussian Kernels as in~\eqref{eq:gaussian-kernel} is to use the \emph{median heuristic} \cite{garreau2018largesampleanalysismedian}.
We choose $\sigma_l^2$ to be the median of the squared distances between all pairs of points in the training set, i.e.,
\begin{equation}
  \sigma_l^2 = \mathrm{median}_{1 \leq i < j \leq N}\left(\|\hat{x}_i-\hat{x}_j\|^2\right),
\end{equation}
where $\hat{x}_1,\ldots,\hat{x}_N$ are the $N$ training samples.
If $N(N-1)/2$ is even, the median is the average of the two middle values in the sorted list of distances.
Better results can be obtained via \emph{grid search}, which consists of evaluating the performance of the estimator for all combinations of \hp in a predefined grid.
This approach is simple to implement and can be very effective for small problems, but it can become computationally expensive for larger problems.
Borrowing a technique from Gaussian processes, we can also try to optimize the \hp to maximize the \emph{log marginal likelihood}, defined as
\begin{equation}
  \log p(\hat{X^+}_N | \hat{X}_N) = - \frac{1}{2}\hat{X^+}_N\T (K_{\hat{X}}^N + N\lambda I_N)^{-1} \hat{X^+}_N -\frac{1}{2}\log |K_{\hat{X}}^N + N\lambda I_N| - \frac{N}{2}\log(2\pi).
\end{equation}

While it is possible to tune the \hp with any method and then set them manually in the estimator, \lucid provides a set of classes, which specialize the \texttt{Tuner} interface, to automate the process.
The \texttt{GridSearchTuner} class implements the grid search method, while the \texttt{MedianHeuristicTuner} class implements the median heuristic.
If the user wants to optimize the log marginal likelihood, they can use the \texttt{LBGSTuner} class, which runs the L-BFGS or L-BFGS-B quasi-Newton optimization algorithms~\cite{book:nonlinear-programming} over the log marginal likelihood function.
Internally, \lucid uses the \texttt{LBFGS++} library~\footnote{https://github.com/yixuan/LBFGSpp} to perform the optimization.
